Open-source landscape for 3D CSEM modelling

### 2020-06-27 manuscript/Manuscript.tex
Lindsey edits to the Intro

% Additional packages to geophysics.cls
\usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]TODO\\
\footnotemark[3]TODO\\
\footnotemark[4]TODO\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: CSEM, Open-Source, 3D modelling, FD, FE.
}

@@

\itodo{
We will have to decide on a Journal. Current suggestions:\\
- GMD: Geoscientific Model Development\\
- SIG: Surveys in Geophysics\\
- SE: Solid Earth\\
- GJI: Geophysical Journal International\\
- GP: Geophysical Prospecting\\
- GEO: Geophysics\\
- CAG: Computers and Geosciences\\
\imycom[DW]{
I think \emph{Geoscientific Model Development} and \emph{Surveys in Geophysics}
were the favourite ones when we discussed it ones. I personally would be very
interested in GMD. What do you think? But also interesting would be the GJI,
because the MT comparison was published there, so it would be in the same
place. I just had a very bad experience with GP, so I am not inclined on that
at all.}
\imycom[RR]{GMD looks very interesting, my only concern is that it might not be
in the scope of EM modelers, but we can probably reach modeleres from other
communities as well.}
\imycom[OCR]{I also find GMD very interesting, in fact it was my first option to
publish the first petgem version (2 years ago). The answer was: "manuscript
outside of journal's scope". In this sense I agree with Raphael. My vote is for
SIG}
}

@@

\imycom[RR]{We need to take care to not mix British English with AE, I'm very
  used to AE, but Dieter started with (at least what I can identify) mainly BE,
  so I tried to follow (mainly s instead of z and ou instead of o in some
  words). I hope there are not too many inconsistencies, as probably Lindsey
  has to fix all these mistakes in the end:-)
\imycom[DW]{Be aware of Canadian English and Australian English :-D -- You are
  of course right. I am usually mostly BE, having done my PhD in Scotland.
  However, I don't think it matters at the moment, because it will depend to
  what journal we submit. With an appropriate spellchecker that should be a
  quick thing then.}}

@@

Controlled-source electromagnetic (CSEM) measurements are a frequently applied
method in various geophysical exploration fields, such as geothermal and
groundwater, oil and gas, mining, civil engineering, or geo-hazards. Modelling
these electromagnetic fields is therefore of great interest to design survey
layouts, to understand the measured data, and for inversion purposes.
Publications regarding 3D modelling in electromagnetic methods started to
appear as early as the 1970's and 1980's. These early publications were
integral equation (IE) methods, having an anomaly embedded within a layered
medium, mostly for loop-loop type transient EM measurements
\citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman} and
magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}.

@@

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of developed numerical
solutions and according publications. The most common applied methods to solve
Maxwell's equation are the IE method \citep{GJI.74.Raiche, RS.02.Hursan,
GEO.06.Zhdanov, GP.10.Tehrani, CAG.16.Kruglyakov, MGS.17.Kruglyakov} and
different variations of the differential equation (DE) method, such as finite
differences (FD) \citep{IEEE.66.Yee, GEO.93.Wang, RSC.94.Mackie, RS.94.Druskin,
GEO.09.Streich, CAG.13.Sommer}, finite elements (FE) \citep{GEO.04.Commer,
GJI.11.Schwarzbach,GEO.12.daSilva, GJI.13.Grayver, GJI.13.Puzyrev,
SEG.16.Zhang}, and finite volumes (FV) \citep{EM.90.Madsen, PIER.01.Clemens,
GP.06.Mulder, ECP.07.Haber, GEO.14.Jahandari}. There are also many different
types of discretisation, where the most common ones are regular grids
(cartesian, rectilinear), mostly using a Yee grid \citep{IEEE.66.Yee} or a
Lebedev grid \citep{CMMP.64.Lebedev}, but also unstructured tetrahedral grids
\citep{SEG.16.Zhang, CAG.17.Cai}, hexagonal meshes \citep{CAG.14.Cai}, or
OcTree meshes\citep{ECP.07.Haber}. Very well written overviews about the
different approaches to 3D EM modelling are given by \cite{SG.05.Avdeev} and
\cite{SG.10.Borner}.

@@

Probably the main reason for the numerous amount of code developments was
already outlined by \cite{SG.05.Avdeev}, who finished his review with the
following statement: «\emph{The most important challenge that faces the EM
community today is to convince software developers to put their 3-D EM forward
and inverse solutions into the public domain, at least after some time. This
would have a strong impact on the whole subject and the developers would
benefit from feedback regarding the real needs of the end-users.}» Some major
exploration service companies, as well as certain research consortia and
research groups, have their own 3D CSEM codes, but how about codes that are
open-source? Academic codes were traditionally often available upon request by
authors. Codes distributed that way were often provided with the request to not
share the code, and they often come without a license attached which means that
they are, legally, not open-source. Also, at times there can be significant
hurdles to install a code.

The meaning of the term open-source evolved rapidly in the last two decades.
Today, open-source generally not only means that a code is available (with a
proper license), but also that the code is hosted online, is under version
control, has the possibility to file issues and make pull requests to fix bugs.
These are important requirements for reproducible research «\emph{to provide
the means by which the reader can verify the validity of the results and make
use of them in further research}» \citep{GEO.17.Broggini}. Well maintained
codes often have continuous integration that includes testing of the code base
and online hosted documentation, often including examples and tutorials. As
such, the term has shifted from purely open-source codes to developments in the
open with increasingly building a global community. Related projects within the
realm of geophysical exploration are e.g., \texttt{pyGIMLi}
\citep{CAG.17.Rucker} or \texttt{Fatiando} \citep{JOSS.18.Uieda}.
\imycom[DW]{Others coming to your mind? An interesting looking one is
  \href{https://github.com/fancompute/ceviche}{github.com/fancompute/ceviche},
  but they are at very high frequencies on nm-distances.
\imycom[RR]{no thoughts about geophysics, but there are many projects for
  related topics such as gmt, gis, mesh-gen}
\imycom[OCR]{No idea from my side}
}

We continue addressing the topic of open-source developments with respect to 3D
CSEM modeling. Even though it comprises still the minority of developments, the
landscape in this field of research changed in the last five years. Before,
there were only closed-source codes owned by companies or  consortia and
individual projects, often without proper documentation, that could be
requested from the authors. We introduce and compare four recent open-source
projects with focus on 3D CSEM modeling. All presented codes are in the Python
ecosystem and use either the FD method on structured grids or the FE method on
unstructured tetrahedral meshes.

Furthermore, we deal with the topic of verification. \cite{GJI.13.Miensopust}
presents a review of two workshops dealing with the verification of
magnetotelluric forward and inversion codes, but we are not aware of any
comparable study or benchmark suite for CSEM data. Hence, we first conduct a
similar exercise for CSEM codes. We want our efforts to help to ensure the
reliability of simulations, not only carried out with the available open-source
codes, but also for individual closed-source projects. Analytical and
semi-analytical solutions only exist for simple halfspace or layered-Earth
models, which served mostly to verify new codes. Beyond them, the only
objective possibility to ensure the accuracy of solutions is by comparing
results from different modellers. If different discretisations and
implementations of Maxwell's equations yield the same result, it gives
confidence in their accuracy.

We simulate EM fields for a layered background model with vertical transverse
isotropy, containing three blocks, and the complex marine Marlim R3D model.
These two model designs as well as the corresponding results from four
different codes provide a benchmark for new codes to be compared to and
validated with. In the following section, we introduce the codes under
consideration. Afterwards, we present the considered benchmark cases in detail
and present the modeling results of our four codes in terms of accuracy and
computational performance. Beyond that, we extensively discuss important points
that control the performance and suitability of our FD and FE codes, including
mainly considerations about the mesh design and the choice of solvers. We
conclude with a summary of our results and a motivation for the EM community to
not only continue to extend the landscape of open-source codes but to also
create a landscape of open-source benchmark models.

>>>>

% Additional packages to geophysics.cls
% \usepackage[UKenglish]{babel}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[T1]{fontenc}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]TODO\\
\footnotemark[3]TODO\\
\footnotemark[4]University of California Berkeley, Department of Statistics\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: CSEM, Open-Source, 3D modelling, FD, FE.
}

@@

\itodo{
We will have to decide on a Journal. Current suggestions:\\
- GMD: Geoscientific Model Development\\
- SIG: Surveys in Geophysics\\
- SE: Solid Earth\\
- GJI: Geophysical Journal International\\
- GP: Geophysical Prospecting\\
- GEO: Geophysics\\
- CAG: Computers and Geosciences\\
\imycom[DW]{
I think \emph{Geoscientific Model Development} and \emph{Surveys in Geophysics}
were the favourite ones when we discussed it ones. I personally would be very
interested in GMD. What do you think? But also interesting would be the GJI,
because the MT comparison was published there, so it would be in the same
place. I just had a very bad experience with GP, so I am not inclined on that
at all.}
\imycom[RR]{GMD looks very interesting, my only concern is that it might not be
in the scope of EM modelers, but we can probably reach modeleres from other
communities as well.}
\imycom[OCR]{I also find GMD very interesting, in fact it was my first option to
publish the first petgem version (2 years ago). The answer was: "manuscript
outside of journal's scope". In this sense I agree with Raphael. My vote is for
SIG}
}
\imycom[LH]{SIG does look like a good option. Given the spirit of the paper, it
would be nice to publish it open-access. Does anyone have a sense of how much that
costs with SIG? I could likely get partial funding towards this. Solid Earth
might be a nice option in that case because it is open-access by default. }

@@

\imycom[RR]{We need to take care to not mix British English with AE, I'm very
  used to AE, but Dieter started with (at least what I can identify) mainly BE,
  so I tried to follow (mainly s instead of z and ou instead of o in some
  words). I hope there are not too many inconsistencies, as probably Lindsey
  has to fix all these mistakes in the end:-)
\imycom[DW]{Be aware of Canadian English and Australian English :-D -- You are
  of course right. I am usually mostly BE, having done my PhD in Scotland.
  However, I don't think it matters at the moment, because it will depend to
  what journal we submit. With an appropriate spellchecker that should be a
  quick thing then.}}
\imycom[LH]{It's true! Canadian English ends up being an interesting hybrid(!)}

@@

Controlled-source electromagnetic (CSEM) measurements are a frequently applied
method in various geophysical exploration fields, such as geothermal,
groundwater, oil and gas, mining, civil engineering, or geo-hazards. Modelling
these electromagnetic fields is therefore of great interest to design survey
layouts, to understand the measured data, and for inversion purposes.
Publications regarding 3D modelling in electromagnetic methods started to
appear as early as the 1970's and 1980's. These early publications were
integral equation (IE) methods, having an anomaly embedded within a layered
medium, mostly for loop-loop type transient EM measurements
\citep{GJI.74.Raiche, GEO.75.Hohmann, GJI.82.Das, GEO.86.Newman} and
magnetotelluric (MT) measurements \citep{GEO.84.Wannamaker}. \cite{Ward1988}
assemble many of these solutions in ``Electromagnetic Theory for Geophysical
Applications,'' which is widely viewed as an authoritative publication on
electromagnetic geophysics.

@@

\imycom[LH]{I added a couple citations here - I think the Haber 2001 is important to include in the FV list (if we want to remove the 2007 one to reduce duplicating authors, I think that is okay as we cite it later)}

The continuous improvement of computing power and the CSEM boom in the early
2000's in the hydrocarbon industry led to a wealth of developed numerical
solutions and according publications. The most common applied methods to solve
Maxwell's equation are the IE method \citep{GJI.74.Raiche, RS.02.Hursan,
GEO.06.Zhdanov, GP.10.Tehrani, CAG.16.Kruglyakov, MGS.17.Kruglyakov} and
different variations of the differential equation (DE) method, such as finite
differences (FD) \citep{IEEE.66.Yee, GEO.93.Wang, RSC.94.Mackie, RS.94.Druskin,
GEO.09.Streich, CAG.13.Sommer}, finite elements (FE) \citep{GEO.04.Commer,
GJI.11.Schwarzbach,GEO.12.daSilva, GJI.13.Grayver, GJI.13.Puzyrev,
SEG.16.Zhang}, and finite volumes (FV) \citep{EM.90.Madsen, Haber2001,PIER.01.Clemens,
GP.06.Mulder, ECP.07.Haber, GEO.14.Jahandari}. There are also many different
types of discretisation, where the most common ones are regular grids
(cartesian, rectilinear), mostly using a Yee grid \citep{IEEE.66.Yee} or a
Lebedev grid \citep{CMMP.64.Lebedev}, but also unstructured tetrahedral grids
\citep{SEG.16.Zhang, CAG.17.Cai}, hexagonal meshes \citep{CAG.14.Cai}, or
OcTree meshes \citep{ECP.07.Haber}. Very well written overviews about the
different approaches to 3D EM modelling are given by \cite{SG.05.Avdeev} and
\cite{SG.10.Borner}.

@@

\imycom[LH]{Here, I think that there
is a point to make about duplication of efforts vs. new methodologies. Often
the development of a new methodology requires that a new implementation be
built from scratch. As a field though, we have matured to a point where we
know what the methods / approaches are, and much of the next stage of
advancements will be more incremental. (We argued this in Oldenburg et al,
2019 - so to self-cite ;) (doi: https://doi.org/10.1080/08123985.2019.1580118,
arxiv: https://arxiv.org/pdf/1902.08245.pdf). See what you think of the
following framing for the next paragraph - we can discuss:

Many of the advancements made in the EM modelling community over the past
decades have required that authors develop new implementation from scratch.
These codes often provided the research group or company with a competitive
advantage for a time, and thus the source codes were often kept internal. In
some cases, executables have been made available for academic purposes upon
request or to sponsoring consortium members. As the field continues to mature,
advancements become more incremental, and particularly in an applied field,
many advancements are driven by new use-cases and applications that were not
considered by the original authors. In his review on EM modelling and
inversion \cite{SG.05.Avdeev} concludes with the
following statement: «\emph{The most important challenge that faces the EM
community today is to convince software developers to put their 3-D EM forward
and inverse solutions into the public domain, at least after some time. This
would have a strong impact on the whole subject and the developers would
benefit from feedback regarding the real needs of the end-users.}» Further,
\cite{Oldenburg2019} argue that an open-source paradigm has the potential to
accelerate multidisciplinary research by encouraging the development of
modular, interoperable tools that are supported by a community of researchers.
}

Probably the main reason for the numerous amount of code developments was
outlined by \cite{SG.05.Avdeev}, who finished his review with the
following statement: «\emph{The most important challenge that faces the EM
community today is to convince software developers to put their 3-D EM forward
and inverse solutions into the public domain, at least after some time. This
would have a strong impact on the whole subject and the developers would
benefit from feedback regarding the real needs of the end-users.}» Some major
exploration service companies, as well as certain research consortia and
research groups, have their own 3D CSEM codes, but how about codes that are
open-source? Academic codes were traditionally often available upon request by
authors. Codes distributed that way are often provided with the request that users not
share the code, and they often come without a license attached which means that
they are, legally, not open-source. Also, at times there can be significant
hurdles to install a code.

\imycom[LH]{Here, I think it is worth discussing the distinction between
open-source: source code available + openly licensed (with an Open Source
Initiative (OSI, https://opensource.org) approved license), open-development:
code is versioned, development and issue tracking happens publicly (e.g. on
GitHub or similar), and open communities: actively soliciting contributions
and pull requests which change the code. and then the push towards open
development / open communities motivates best practices including continuous
integration, documentation, etc. See what you think of my suggested adaptation
below. There might be a few citations we can add for an open model of development / open community project. I will do some digging. Again, happy to discuss.

In many domains of science, it is becoming more common for researchers to
release source-code with an open license that allows use and modification (
e.g. see the Open Source Initiative approved licenses:
\href{https://opensource.org/licenses}{https://opensource.org/licenses}). This
is an important step for improving reproducibility of research: «\emph{to
provide the means by which the reader can verify the validity of the results
and make use of them in further research}» \citep{GEO.17.Broggini}. Going a
step beyond releasing open-source software, many groups adopt an open model of
development where code is hosted and versioned in an online repository,
all changes are public, and users can engage by submitting and track issues.
Community oriented projects further engagement by encouraging pull-requests,
which are suggested changes to the code. Successful, well-maintained projects
often have unit-testing and continuous integration that runs those tests with
any changes to the code. Additionally, documentation that includes examples
and tutorials is an important component for on-boarding new users and
contributors. As a result, in many areas of the geosciences, we are seeing a
shift away from a one-way distribution of code towards building global
communities around open projects. Related projects within the
realm of geophysical exploration are e.g., \texttt{pyGIMLi}
\citep{CAG.17.Rucker} or \texttt{Fatiando} \citep{JOSS.18.Uieda}
}

The meaning of the term open-source evolved rapidly in the last two decades.
Today, open-source generally not only means that a code is available (with a
proper license), but also that the code is hosted online, is under version
control, has the possibility to file issues and make pull requests to fix bugs.
These are important requirements for reproducible research «\emph{to provide
the means by which the reader can verify the validity of the results and make
use of them in further research}» \citep{GEO.17.Broggini}. Well maintained
codes often have continuous integration that includes testing of the code base
and online hosted documentation, often including examples and tutorials. As
such, the term has shifted from purely open-source codes to developments in the
open with increasingly building a global community. Related projects within the
realm of geophysical exploration are e.g., \texttt{pyGIMLi}
\citep{CAG.17.Rucker} or \texttt{Fatiando} \citep{JOSS.18.Uieda}.
\imycom[DW]{Others coming to your mind? An interesting looking one is
  \href{https://github.com/fancompute/ceviche}{github.com/fancompute/ceviche},
  but they are at very high frequencies on nm-distances.
\imycom[RR]{no thoughts about geophysics, but there are many projects for
  related topics such as gmt, gis, mesh-gen}
\imycom[OCR]{No idea from my side}
\imycom[LH]{Could we cite: https://github.com/softwareunderground/awesome-open-geoscience ?. Maybe we could ask them to put it on zenodo so it has a doi?? Also Madagascar in seismic?}
}

In this paper, we discuss the topic of open-source developments with respect to 3D
CSEM modeling. Even though it comprises still the minority of developments, the
landscape in this field of research changed in the last five years. Previously,
there were only closed-source codes owned by companies or  consortia and
individual projects, often without proper documentation, that could be
requested from the authors. We introduce and compare four recent open-source
projects with focus on 3D CSEM modeling. All presented codes are in the Python
ecosystem and use either the FV method on structured grids or the FE method on
unstructured tetrahedral meshes.

Furthermore, we deal with the topic of verification. \cite{GJI.13.Miensopust}
presents a review of two workshops dealing with the verification of
magnetotelluric forward and inversion codes, but we are not aware of any
comparable study or benchmark suite for CSEM data. Hence, we first conduct a
similar exercise for CSEM codes. We want our efforts to help to ensure the
reliability of simulations, not only carried out with the available open-source
codes, but also for individual closed-source projects. Analytical and
semi-analytical solutions only exist for simple halfspace or layered-Earth
models, which served mostly to verify new codes. Beyond these simple models, the only
objective possibility to ensure the accuracy of solutions is by comparing
results from different modellers. If different discretisations and
implementations of Maxwell's equations yield the same result, it gives
confidence in their accuracy.

We simulate EM fields for a layered background model with vertical transverse
isotropy, containing three blocks, as well as for the complex marine Marlim R3D model.
These two model designs as well as the corresponding results from four
different codes provide a benchmark for new codes to be compared to and
validated with. In the following section, we introduce the codes under
consideration. Afterwards, we present the considered benchmark cases in detail
and present the modeling results of our four codes in terms of accuracy and
computational performance. Beyond that, we extensively discuss important points
that control the performance and suitability of our FV and FE codes, including
mainly considerations about the mesh design and the choice of solvers. We
conclude with a summary of our results and a motivation for the EM community to
not only continue to extend the landscape of open-source codes but to also
create a landscape of open-source benchmark models.
